Stability Analysis of Recurrent Neural Networks \\ using Ordinary Differential Equations

### 2020-07-11 p5/stability_rnn.tex
Third revision.
Fixed a typo.
A better filler.
Updated with LTSM architecture

\documentclass{article}
\usepackage{amsmath, amssymb, amsfonts, amsthm}
\numberwithin{equation}{section}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem*{rem}{Remark}
\begin{document}
\title{Stability Analysis of Recurrent Neural Networks \\
    using Ordinary Differential Equations}
\author{Dan Will}
\maketitle
\section{Introduction}\label{s1}
Recurrent neural networks feed their output to their input. This feature
makes them look like algorithms to get numerical solutions of ordinary
differential equations. It is well known that training neural networks is
rife with vanishing and exploding gradients\cite{gereon2018hands}. This
phenomenon is closely related to the behaviour of the numerical algorithms
to solve ordinary differential equations, a topic that has been studied
intensively for several decades. We use this analogy to derive criteria for
the stability of recurrent neural networks.

@@

The oldest method to solve equation \eqref{s3e1} numerically was devised 
by Euler in the eighteenth century. It follows straight from the definition
of the first derivative of a function. If we know the value of $y$ at $x_0$,
we can approximate the value of $y$ at a neighbouring point $x_0 + h$ as
\begin{equation}\label{s3e2}
y(x_0 + h) = y(x_0) + h\left[\frac{dy}{dx}\right]_{x_0}.
\end{equation}
But the value of the derivative of $y$ with respect to $x$ is given by
the differential equation itself. Thus,
\begin{equation}\label{s3e3}
y(x_0 + h) = y(x_0) + h f(x_0, y(x_0)).
\end{equation}
In general, $n$ steps beyond the first, one can write
\begin{equation}\label{s3e4}
y(x_n + h) = y(x_{n}) + h f(x_{n}, y(x_{n})).
\end{equation}
Euler's algorithm is not the best numerical algorithm to solve ODE's but
it provides the germ for the rest.

\section{RNN with a Single Layer}\label{s4}
Recurrent neural networks, abbreviated as RNN, feed a part of their output
to their input. The output of a step $n$ thus depends on both the input
at that step and the output of the previous step. This makes them similar
to the ordinary differential equations described in the previous section.
Suppose we have a single layer neural network that is updated using the
following rule:
\begin{equation}\label{s4e1}
y_{n+1} = \sigma(w_x x_{n} + w_y y_n),
\end{equation}
where $w_x$ and $w_y$ are weights and $\sigma$ is an activation function. 
If we now identify the correspondence
\begin{eqnarray}
y_{n+1} &\rightarrow& y(x_n + h) \\
y_{n} &\rightarrow& y(x_n) \\
\sigma(w_x x_n + w_y y_n) &\rightarrow& y(x_n) + hf(x_n, y(x_n))
\end{eqnarray}
then we can spot a striking resemblance between Euler's algorithm and an
RNN. The idea that an RNN corresponds to an ODE and therefore the stability
of an RNN can be studied using the stability of the underlying ODE is the
topic of this article.

@@

Consider the initial value problem of equation \eqref{s3e1}. The general
$k$-step method to get its approximate numerical solution is defined by
the formula\cite{dahlquist1963special}
\begin{equation}\label{s9e1}
\sum_{i=0}^k \alpha_i x_{n+i} = h\sum_{i=0}^k \beta_i f_{n+i},
\end{equation}
where $\alpha_i, \beta_i$ are real constants.
\begin{defn}\label{s9d1}
The $k$-step method defined above is said to be A-stable
\cite{dahlquist1963special} if all solutions of \eqref{s9e1} tend to zero 
as $n \rightarrow \infty$ when the method is applied to the equation
\[
    \frac{dy}{dx} = qx
\]
where $q \in \mathbb{C}$ with $\Re(q) < 0$ and a fixed $h$ in equation
\eqref{s9d1}.
\end{defn}

@@

The GRU described in section \ref{s8} has $\alpha_{i, j} = 0$ because the
hidden layers are not connected to each other. The Burrage-Butcher tensor 
is thus $-\beta_1 \beta_1^T$, which is \emph{not} positive semidefinite 
unless $\beta_1 = 0$.  Thus the GRU is not a stable algorithm, an 
observation also corroborated by elsewhere\cite{kanai2017preventing}.
\bibliographystyle{plain}
\bibliography{stability_rnn}
\end{document}

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, amsfonts, amsthm}
\numberwithin{equation}{section}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem*{rem}{Remark}
\begin{document}
\title{Stability Analysis of Recurrent Neural Networks \\
    using Ordinary Differential Equations}
\author{Author's name}
\maketitle
\section{Introduction}\label{s1}
Recurrent neural networks feed their output to their input. This feature
makes them look like algorithms to get numerical solutions of ordinary
differential equations. It is well known that training neural networks is
rife with vanishing and exploding gradients\cite{gereon2018hands}. This
phenomenon is closely related to the behaviour of the numerical algorithms
to solve ordinary differential equations, a topic that has been studied
intensively for several decades. We use this analogy to bank upon the 
vast research about ordinary differential equations to derive criteria 
for the stability of recurrent neural networks.

@@

The oldest method to solve equation \eqref{s3e1} numerically was devised 
by Euler in the eighteenth century. It follows straight from the definition
of the first derivative of a function. If we know the value of $y$ at $x_0$,
we can approximate the value of $y$ at a neighbouring point $x_0 + h$ as
\begin{equation}\label{s3e2}
y(x_0 + h) = y(x_0) + h\left[\frac{dy}{dx}\right]_{x_0}.
\end{equation}
The square bracket around the derivative and the subscript $x_0$ to the
right bracket indicates that the derivative is to be evaluated at $x = x_0$.
The value of the derivative of $y$ with respect to $x$ is given by
the differential equation itself. Thus,
\begin{equation}\label{s3e3}
y(x_0 + h) = y(x_0) + h f(x_0, y(x_0)).
\end{equation}
In general, $n$ steps beyond the first, one can write
\begin{equation}\label{s3e4}
y(x_n + h) = y(x_{n}) + h f(x_{n}, y(x_{n})).
\end{equation}
Euler's algorithm is not the best numerical algorithm to solve ODE's but
it provides the germ for the rest.

\section{RNN with a Single Layer}\label{s4}
Recurrent neural networks, abbreviated as RNN, feed a part of their output
to their input. The output of a step $n$ thus depends on both the input
at that step and the output of the previous step. This makes them similar
to the ordinary differential equations described in the previous section.
Suppose we have a single layer neural network that is updated using the
following rule:
\begin{equation}\label{s4e1}
y_{n+1} = \sigma(w_x x_{n} + w_y y_n),
\end{equation}
where $w_x$ and $w_y$ are weights and $\sigma$ is an activation function. 
If we now identify the correspondence
\begin{eqnarray}
y_{n+1} &\rightarrow& y(x_n + h) \\
y_{n} &\rightarrow& y(x_n) \\
\sigma(w_x x_n + w_y y_n) &\rightarrow& y(x_n) + hf(x_n, y(x_n))
\end{eqnarray}
then we can spot a resemblance between Euler's algorithm and an
RNN. The idea that an RNN corresponds to an ODE and therefore the stability
of an RNN can be studied using the stability of the underlying ODE is the
topic of this article.

@@

Consider the initial value problem of equation \eqref{s3e1}. The general
$k$-step method to get its approximate numerical solution is defined by
the formula\cite{dahlquist1963special}
\begin{equation}\label{s9e1}
\sum_{i=0}^k \alpha_i x_{n+i} = h\sum_{i=0}^k \beta_i f_{n+i},
\end{equation}
where $\alpha_i, \beta_i$ are real constants.
\begin{defn}\label{s9d1}
The $k$-step method defined above is said to be A-stable
\cite{dahlquist1963special} if all solutions of \eqref{s9e1} tend to zero 
as $n \rightarrow \infty$ when the method is applied to the equation
\[
    \frac{dy}{dx} = qx
\]
where $q \in \mathbb{C}$ with $\Re(q) < 0$ and a fixed $h$ in equation
\eqref{s9d1}. Here $n$ is the number of steps in the numerical algorithm.
\end{defn}

@@

The GRU described in section \ref{s8} has $\alpha_{i, j} = 0$ because the
hidden layers are not connected to each other. The Burrage-Butcher tensor 
is thus $-\beta_1 \beta_1^T$, which is \emph{not} positive semidefinite 
unless $\beta_1 = 0$.  Thus the GRU is not a stable algorithm, an 
observation also corroborated by elsewhere\cite{kanai2017preventing}.

\section{GRU with an Attention Layer}\label{s10}
We will now analyze GRU with attention layer. We assume that the 
architecture of the network is same as shown in figure 4 of the paper
\cite{ran2019lstm} by Xiangdong and others, except that the LTSM is 
replaced with a GRU. There are $n$ such GRU's, each with an output
$h^i_l \in \mathbb{R}^s$. The combined output of all of them is the $s
\times n$ matrix $Y_l$ defined by
\begin{equation}\label{s10e1}
H_l = [h^1_l, \ldots, h^n_l].
\end{equation}
The matrix $H_l$ is scaled with an $n \times 1$  weight matrix $W_H$. 
Similarly, if $A$ is an $s \times n$ embedding matrix and $W_v$ is the
corresponding weight matrix, then
\begin{equation}\label{s10e2}
M_l = \tanh\begin{pmatrix} H_lW_H \\ AW_v \end{pmatrix}
\end{equation}
is a member of $\mathbb{R}^{2s}$. If $W= [w_1, \ldots, w_n]$ is a $(2s) 
\times n$ weight matrix then
\begin{equation}\label{s10e3}
a_i = \frac{\exp(w_i^T M_l)}{\sum_{j=1}^n \exp(w_j^T M_l)}
\end{equation}
are the $n$ components of the vector $a = (a_1, \ldots, a_n)
\in \mathbb{R}^n$. Finally, the product
\begin{equation}\label{s10e4}
Y_l = H_l a
\end{equation}
is a member of $\mathbb{R}^s$. Note that the output of the individual
GRUs is also an $s$-dimensional vector.

Equations \eqref{s10e1} to \eqref{s10e4} where $Y_l$ is gived by 
equations \eqref{s8e1} to \eqref{s8e5} defined GRU with an attention layer.
We will analyse the stability of this architecture in this section.

Equation \eqref{s10e4} tells that $Y_l$ depends $H_l$ but not the $H_{l-1}$.
Thus, it is not a recurrence relation. $Y_l$ is not computed from the
`present' $H_l$, not a `previous' one. Therefore, this attention layer 
does not play a role in determining the stability of the network in the
architecture shown in figure 4 of the paper \cite{ran2019lstm} by 
Xiangdong and others.

\bibliographystyle{plain}
\bibliography{stability_rnn}
\end{document}
